Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-11 manuscript.tex
latest

The classic lottery model is generalized to accomodate arbitrary densities. We use the model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal theory of plant primary strategies. We evaluate the density-dependent lottery model as a general-purpose model of density-dependent selection in the spirit of MacArthur's $r$/$K$ theory.

@@

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2012}.

However, a genotype's selective advantage

 can have very different effects when operating on different types of traits, and evolutionary changes in population can lead to complicated ecological responses.

@@

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose. The lottery model gets its name from the way in which it represents competition: the growth of each type is determined by the number of contestants that it produces relative to the other types in the population (akin to a lottery),  weighted by a coefficient for each type representing its competitive ability. This is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}. Here we extend the classic lottery model to correctly account for low density behavior.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. Each generation, propagules are produced by adults which are dispersed at random over the unoccupied territories. Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then death occurs. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').} 
\end{figure}

@@

In addition to propagule birth and competition, occupied territories become unoccupied due to mortality. For the majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
We will introduce a different mortality model when we consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), which will also affect competing juveniles. However, in the absence of severe disturbances, restricting mortality to adults (setting aside the deaths implicit in territorial contest) is a reasonable approximation  \label{eq:delttot}

@@

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$. In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}. Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive. Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals. However, for fishes, the analogous ``periodic'' strategy is characterized by enormous brood sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$). The evolution of $b$ in ideal environments will be discussed further in the Discussion.

>>>>

Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients. This excludes fundamental ecological factors such as dynamic population size or density-dependence from our most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory. Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests. To do so, we generalize the classic lottery model of territorial acquisition, which has primarily been used for studying species co-existence questions, to accommodate arbitrary densities. We use this density-dependent lottery model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal scheme. We revisit previous concepts of density-dependent selection, including $r$ and $K$ selection, and argue that our model distinguishes between different aspects of fitness in a more natural and intuitive manner.

@@

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2012}. However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.

@@

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose. The lottery model gets its name from the way in which it represents competition: the growth of each type is determined by the number of contestants that it produces relative to the other types in the population (akin to a lottery),  weighted by a coefficient for each type representing its competitive ability. This is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}. Here we analytically extend the classic lottery model to correctly account for low density behavior.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. Each generation, propagules are produced by adults which are dispersed at random over the unoccupied territories. Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then occupied territories are freed up by adult mortality. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').} 
\end{figure}

@@

In addition to propagule birth and competition, occupied territories become unoccupied due to mortality. For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
This is reasonable approximation in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).

@@

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$. In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}. Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive. Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals. However, for fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$). The evolution of $b$ in ideal environments will be discussed further in the Discussion.
